Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-07-28 jso.qmd
consolidate the definition in PP and squintability

@@

\begin{thm}\label{thm:my_theorem}
This is a theorem.
\end{thm}
Theorem \ref{thm:my_theorem} is important.
-->
```
Projection Pursuit (PP) (@kr69, @FT74) is a dimension reduction
technique aimed at identifying informative linear projections of data.
The method involves optimising an objective function known as the PP
index (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), which defines the
criteria for what constitutes interesting or informative projections.
Given high-dimensional data $X \in \mathbf{R}^{n\times p}$ and an index
function $f(\cdot)$, PP finds the orthonormal basis
$A \in \mathbf{R}^{p \times d}$ that maximises the index value of the
projection, $Y = XA$:

$$
\underset{A}{\max } \quad f(XA) \quad \text{subject to} \quad A'A = I
$$ {#eq-optimization}

@@

```{=tex}
\begin{defn}\label{def:proj-dist}
Let $A \in \mathbf{R}^{p \times d}$ be the d-dimensional projection basis and
$A^*$ be the optimal basis that achieves the maximum index value for a given
PP problem defined by the shape to find, data dimension,
and the index function. The projection distance $d(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$d(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\  \rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
```
We now focus our attention to how the index value $f(XA)$ changes with
respect to the projection distance $d(A, A^*)$, over the course of the
JSO.

```{=tex}
\begin{defn}\label{def:squintability}
Let $g$ be the function that maps the projection distance, $d = d(A, A^*)$, 
to the index value $f(XA)$, such that $g(d(A, A^*)) = f(XA)$.
The squintability of $f(\cdot)$, denoted $\varsigma(f)$, is defined as 
$$\varsigma(f) = c \times \max_{d} \Delta g(d) \times \arg \max_{d}  \Delta g(d)$$ 
where $c$ is a constant scaling factor, $\max_A \Delta g(x)$ represents the 
maximum change in the index value along the curve $g$, and $\arg \max_{d} \Delta g(x)$ represents the projection distance at which this largest gradient is attained.

\end{defn}
```
Definition \ref{def:squintability} is ...

@@

p1 <- aaa |> 
  ggplot() + 
  geom_point(aes(x = V1, y = V2), size = 0.05) + 
  geom_text(data = aaa |> select(-V1, -V2) |> unique(), 
            aes(label = round(index_val,3)), x =0, y = 3.7, 
            size = 3) + 
  facet_grid(d ~ id) + 
  xlim(-4, 4) + ylim(-4, 4) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())

bbb <- pipe_better |> 
  group_by(dim) |> 
  arrange(index_val) |> 
  filter(row_number() %in% c(1, seq(5, 50, 5))) |>  
  mutate(id = factor(
    paste0(c(0, seq(5, 50, 5)) / 50 * 100, "th"),
    levels = paste0(c(0, seq(5, 50, 5)) / 50 * 100, "th")), 
    d= 4 + 2 * as.numeric(dim)) |> 
  unnest(proj)

p2 <- bbb |> 
  ggplot() + 
  geom_point(aes(x = V1, y = V2), size = 0.05) + 
  geom_text(data = bbb |> select(-V1, -V2) |> unique(), 
            aes(label = round(index_val,3)), x =0, y = 3.7, 
            size = 3) + 
  facet_grid(d ~ id) + 
  xlim(-4, 4) + ylim(-4, 4) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())

>>>>

editor_options: 
  chunk_output_type: console

@@

\begin{thm}\label{thm:my_theorem}
This is a theorem.
\end{thm}
Theorem \ref{thm:my_theorem} is important.
-->
```
Projection Pursuit (PP) (@kr69, @FT74) is a dimension reduction
technique aimed at identifying informative linear projections of data.
The method involves optimising an objective function known as the PP
index (e.g. @hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020), which defines the
criteria for what constitutes interesting or informative projections.
Let $X \in \mathbb{R}^{n\times p}$ be the data matrix, $A \in\mathbb{R}^{p \times d}$ be an orthonormal matrix, where $A$ belongs to the Stiefel manifold $\mathcal{A} = V_d(\mathbb{R}^p)$. The projection $Y = XA$ is a linear transformation that maps data from a $p$-dimensional space into a $d$-dimensional space. The index function $f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}$ is a scalar function that measures an interesting aspect of the projected data, such as deviation from normality, presence of clusters, non-linear structure, or other features of interest. For a fixed data, PP finds the orthonormal basis $A$ that maximises the index value of the projection, $Y = XA$:

$$
\underset{A \in \mathcal{A}}{\max } \quad f(XA) \quad \text{subject to} \quad A'A = I_d
$$ {#eq-optimization}

@@

```{=tex}
\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbf{R}^{p \times d}$ be the d-dimensional projection basis and
$A^*$ be the optimal basis that achieves the maximum index value for a given
PP problem defined by the shape to find, data dimension,
and the index function. The projection distance $d(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$d(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\  \rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
```
We now focus our attention to how the index value $f(XA)$ changes with
respect to the projection distance $d(A, A^*)$, over the course of the
JSO.

<!-- Squintability is defined as how rapidly and significantly the index value, $f(XA)$ response to the decrease of projection distance to optimal, $d(A, A^*)$. -->

```{=tex}
\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance, $d = d(A, A^*)$, 
to the index value $f(XA)$, such that $g(d(A, A^*)) = f(XA)$. The squintability of the index function $f(\cdot)$, is defined as 
$$\varsigma(f) = -c \times \max_{d} g'(d) \times \arg \max_{d} g'(d)$$ 
where $c$ is a constant scaling factor, $-\max_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \max_{d} g'(d)$ represents the projection distance at which this largest gradient is attained.

\end{defn}
```

<!-- Definition \ref{def:squintability} is ... -->

@@

p1 <- aaa |> 
  ggplot() + 
  geom_point(aes(x = V1, y = V2), size = 0.05) + 
  geom_text(data = aaa |> select(-V1, -V2) |> unique(), 
            aes(label = round(index_val,3)), x = 0, y = 3.7, 
            size = 3) + 
  facet_grid(d ~ id) + 
  xlim(-4, 4) + ylim(-4, 4) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())

bbb <- pipe_better |> 
  group_by(dim) |> 
  arrange(index_val) |> 
  filter(row_number() %in% c(1, seq(5, 50, 5))) |>  
  mutate(id = factor(
    paste0(c(0, seq(5, 50, 5)) / 50 * 100, "th"),
    levels = paste0(c(0, seq(5, 50, 5)) / 50 * 100, "th")), 
    d = 4 + 2 * as.numeric(dim)) |> 
  unnest(proj)

p2 <- bbb |> 
  ggplot() + 
  geom_point(aes(x = V1, y = V2), size = 0.05) + 
  geom_text(data = bbb |> select(-V1, -V2) |> unique(), 
            aes(label = round(index_val,3)), x = 0, y = 3.7, 
            size = 3) + 
  facet_grid(d ~ id) + 
  xlim(-4, 4) + ylim(-4, 4) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())
